---
title: "Building MCMC-based models of SI for SARS-Cov-2"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes 
    theme: united
    toc: yes 
---

# Tables with primary and secondary contact events

This tables are formed to help produce samples from the SI distribution
for SARS-Cov-2.  The source is [Du et al. (2020)](https://wwwnc.cdc.gov/eid/article/26/6/20-0357_article).
```{r lktab}
library(sars2pack)
dim(pos_si_93cn)
head(pos_si_93cn)
```
`pos_si_93cn` is a subset of 468 records with positive serial interval.
```{r lk2}
dim(cens_si_93cn)
cens_si_93cn[10:15,]
```
`cens_si_93cn` constitutes a modification to the serial interval data to
avoid introducing negative serial intervals, retaining all 468
records.  Any record in which the
symptom onset of the secondary case is prior to that of the index case
is reformulated with the index case time regarded as interval-censored
with onset date known only to lie in an interval of length 28 days, ending
one day before the onset date reported for the secondary case.

## MCMC estimation of the SI distribution

### Using the censored data representation

These entities were precomputed Apr 24 2020.
```{r doco,eval=FALSE}
library(coarseDataTools)
library(EpiEstim)
censans_W = coarseDataTools::dic.fit.mcmc(dat = cens_si_93cn, dist = "W", optim.method="SANN",
   seed=1234)
si_cens_W = coarse2estim(censans_W, thin=10)
save(si_cens_W, file="si_cens_W.rda")
censans_L = coarseDataTools::dic.fit.mcmc(dat = cens_si_93cn, dist = "L", optim.method="SANN",
   seed=1234)
si_cens_L = coarse2estim(censans_L, thin=10)
save(si_cens_L, file="si_cens_L.rda")
censans_G = coarseDataTools::dic.fit.mcmc(dat = cens_si_93cn, dist = "G", optim.method="SANN",
   seed=1234)
si_cens_G = coarse2estim(censans_G, thin=10)
save(si_cens_G, file="si_cens_G.rda")
```

On the left we trace the pointwise means of the posterior distribution of SI(t).
On the right we give the posterior density of SI(4) for three
parametric models used to form the censored data likelihoods.

```{r lkpl, fig.height=5, echo=FALSE}
ln = si_cens_L
w = si_cens_W
g = si_cens_G
par(mfrow=c(1,2))
plot(1:40,apply(ln[[1]], 1, mean)[1:40], type="l", lwd=2, xlab="days from origin", ylab="density",
    col="darkgray", main="SI(t), posterior mean (pointwise)")
lines(1:40, apply(w[[1]], 1, mean)[1:40], type="l", lty=2, lwd=2, col="darkgray")
lines(1:40, apply(g[[1]], 1, mean)[1:40], type="l", lty=3, lwd=2, col="darkgray")
legend(20, .14, lty=1:3, legend=c("lognormal", "Weibull", "Gamma"), col="darkgray", bty="n", lwd=2)

plot(density(ln[[1]][4,]), xlim=c(.10,.18), lwd=2, col="darkgray", main="Posterior mass at day 4", xlab="MCMC mass")
lines(density(w[[1]][4,]), lwd=2, col="darkgray", lty=2)
lines(density(g[[1]][4,]), lwd=2, col="darkgray", lty=3)
```

### Using only positive values for transmission time

```{r dopos, cache=TRUE, echo=FALSE, message=FALSE}
posans_W = coarseDataTools::dic.fit.mcmc(dat = pos_si_93cn, dist = "W", optim.method="SANN",
   seed=1234, verbose=0)
si_pos_W = coarse2estim(posans_W, thin=10)
save(si_pos_W, file="si_pos_W.rda")
posans_L = coarseDataTools::dic.fit.mcmc(dat = pos_si_93cn, dist = "L", optim.method="SANN",
   seed=1234, verbose=0)
si_pos_L = coarse2estim(posans_L, thin=10)
save(si_pos_L, file="si_pos_L.rda")
posans_G = coarseDataTools::dic.fit.mcmc(dat = pos_si_93cn, dist = "G", optim.method="SANN",
   seed=1234, verbose=0)
si_pos_G = coarse2estim(posans_G, thin=10)
save(si_pos_G, file="si_pos_G.rda")
```

```{r lkpl2, fig.height=5, echo=FALSE}
ln = si_pos_L
w = si_pos_W
g = si_pos_G
par(mfrow=c(1,2))
plot(1:40,apply(ln[[1]], 1, mean)[1:40], type="l", lwd=2, xlab="days from origin", ylab="density",
    col="darkgray", main="SI(t), posterior mean (pointwise)")
lines(1:40, apply(w[[1]], 1, mean)[1:40], type="l", lty=2, lwd=2, col="darkgray")
lines(1:40, apply(g[[1]], 1, mean)[1:40], type="l", lty=3, lwd=2, col="darkgray")
legend(20, .14, lty=1:3, legend=c("lognormal", "Weibull", "Gamma"), col="darkgray", bty="n", lwd=2)

plot(density(ln[[1]][4,]), xlim=c(.10,.18), lwd=2, col="darkgray", main="Posterior mass at day 4", xlab="MCMC mass")
lines(density(w[[1]][4,]), lwd=2, col="darkgray", lty=2)
lines(density(g[[1]][4,]), lwd=2, col="darkgray", lty=3)
```


```{r getdat, cache=TRUE}
suppressWarnings({dat = enriched_jhu_data()})
cusa = cumulative_events_ejhu(dat, eventtype="confirmed",
      alpha3="USA")
cusa
cusa = trim_from(cusa, "2020-03-01")
iusa = form_incident_events(cusa)
```
```{r fit1, fig.height=6}
wcens = EpiEstim::estimate_R(iusa$count, method="si_from_sample", si_sample=si_cens_W$si_sample, 
     config=make_config(list(n2=50, seed=1)))
nn = plot(wcens)
```
```{r fit2, fig.height=6}
lncens = EpiEstim::estimate_R(iusa$count, method="si_from_sample", si_sample=si_cens_L$si_sample, 
     config=make_config(list(n2=50, seed=1)))
nn = plot(lncens)
```



```
sann_pos = .Last.value
names(sann_pos)
str(sann_pos)
?estimate_R
si_sam = coarse2estim(sann_pos, thin=10)
names(si_sam)
EpiEstim::estimate_R(iusa$count, method="si_from_sample", si_sample=si_sam, config=make_config(list(n2=50,
overall_seed=1))) -> nreu
EpiEstim::estimate_R(iusa$count, method="si_from_sample", si_sample=si_sam, config=make_config(list(n2=50,
seed=1))) -> nreu
EpiEstim::estimate_R(iusa$count, method="si_from_sample", si_sample=si_sam$si_sample, config=make_config(list(n2=50,
seed=1))) -> nreu
```
