---
title: "sars2pack demonstration: working with new data"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{sars2pack: working with new data}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

```{r setup,echo=FALSE}
suppressMessages({
suppressPackageStartupMessages({
library(BiocStyle)
library(sars2pack)
library(dplyr)
library(tibble)
library(lubridate)
})
})
```

# New data setup


```{r newdates}
suppressMessages({
suppressPackageStartupMessages({
library(sars2pack)
library(lubridate)
})
})
datevec = as_date(mdy(
c("1/22/20", "1/23/20", "1/24/20", "1/25/20", "1/26/20", "1/27/20", 
"1/28/20", "1/29/20", "1/30/20", "1/31/20", "2/1/20", "2/2/20", 
"2/3/20", "2/4/20", "2/5/20", "2/6/20", "2/7/20", "2/8/20", "2/9/20", 
"2/10/20", "2/11/20", "2/12/20", "2/13/20", "2/14/20", "2/15/20", "2/16/20", 
"2/17/20", "2/18/20", "2/19/20", "2/20/20", "2/21/20", "2/22/20", 
"2/23/20", "2/24/20", "2/25/20", "2/26/20", "2/27/20", "2/28/20", 
"2/29/20", "3/1/20", "3/2/20", "3/3/20", "3/4/20", "3/5/20", "3/6/20", 
"3/7/20", "3/8/20", "3/9/20", "3/10/20", "3/11/20", "3/12/20", "3/13/20", 
"3/14/20", "3/15/20", "3/16/20", "3/17/20", "3/18/20")))
cumevents = c(1, 1, 2, 2, 5, 5, 5, 5, 5, 7, 8, 8, 11, 11, 11, 11, 11, 
11, 11, 11, 12, 12, 13, 13, 13, 13, 13, 13, 13, 13, 15, 15, 
15, 51, 51, 57, 58, 60, 68, 74, 98, 118, 149, 217, 262, 402, 
518, 583, 959, 1281, 1663, 2179, 2727, 3499, 4632, 6421, 7783)
names(cumevents) = datevec
plot(cumevents~as_date(ymd(names(cumevents))), main="US as of 3/18/2020 from Mallery")
```

# Basic estimation and plotting

Now we work with the R0 package functions.  The plotting
titles are hard to manage, and the code from package R0
will need to be dissected to improve annotation of figures.
```{r dor0, fig.height=7}
mGT <- generation.time("gamma", c(5.8,.95)) # from DOI 10.7326/M20-0504
usdat = trim_leading_zeros(diff(cumevents))
est1 = estimate.R( usdat, GT = mGT, t = as_date(ymd(names(usdat))),
  begin=1L, end=as.integer(length(usdat)), methods=c("EG")) # , "ML", "TD"))
est1
par(mfrow=c(1,2))
plot2(est1)
plotfit2(est1)
```

# Sensitivity analysis

The R0 package includes sensitivity analysis support.
We need to specify features of the generation time
models, and ranges of R0 are produced.

```{r lksa}
sa1 = sa.GT(incid=usdat, GT.type="gamma", GT.mean=seq(3,7,1), GT.sd.range=2,
        begin=1L, end=as.integer(length(usdat)), est.method="EG")
sa1 = apply(sa1[,-1], 2, as.numeric)
plot(x=sa1[,"GT.Mean"], xlab="mean GT (days)", y=sa1[,"R"], ylim=c(2, 7), ylab="R0 (95pct CI)",
          type="p", pch=19, col="black", main="Sensitivity of R0 to mean GT")
arrows(x0=sa1[,"GT.Mean"], y0=sa1[,"CI.lower"], 
            y1=sa1[,"CI.upper"], angle=90, code=3, col="black", length=0.05)
```


