---
title: "Building MCMC-based models of SI for SARS-Cov-2"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{"Building MCMC-based models of SI for SARS-Cov-2"}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes 
    theme: united
    toc: yes 
---

```{r setup, echo=FALSE, results="hide"}
suppressPackageStartupMessages({
library(sars2pack)
library(EpiEstim)
})
```

# Overview

The serial interval is defined as the time elapsed between
onset of symptoms in an index case, and onset of symptoms
in a secondary case infected through contact with the index case.


The distribution of the serial interval plays a role in modeling time-dependent
transmission rates.  Specifically, following Cori et al. (2013), the 
instantaneous reproduction number, that we will denote R[t], is defined
as
\[
R[t] = I[t]/\Sigma_{s=1}^t I[t-s]w[s]
\]
where $I[t]$ denotes incidence of infection in period $t$ and $w[s]$ is the
"infectivity function" which is "approximated by the distribution of the
serial interval" (p1506).


# Tables with primary and secondary contact events

This tables are formed to help produce samples from the SI distribution
for SARS-Cov-2.  The source is [Du et al. (2020)](https://wwwnc.cdc.gov/eid/article/26/6/20-0357_article).
```{r lktab}
library(sars2pack)
dim(pos_si_93cn)
head(pos_si_93cn)
```
`pos_si_93cn` is a subset of 468 records with positive serial interval.
```{r lk2}
dim(cens_si_93cn)
cens_si_93cn[10:15,]
```
`cens_si_93cn` constitutes a modification to the serial interval data to
avoid introducing negative serial intervals, retaining all 468
records.  Any record in which the
symptom onset of the secondary case is prior to that of the index case
is reformulated with the index case time regarded as interval-censored
with onset date known only to lie in an interval of length 28 days, ending
one day before the onset date reported for the secondary case.

## MCMC estimation of the SI distribution

The crucial resource here is the approach to
fitting models to (potentially) doubly interval-censored
event time data as 
developed by [N. Reich et al., Statistics in Medicine 2009](http://doi.wiley.com/10.1002/sim.3659).

### Using the censored data representation

These entities were precomputed Apr 24 2020.
```{r doco,eval=FALSE}
library(coarseDataTools)
library(EpiEstim)
censans_W = coarseDataTools::dic.fit.mcmc(dat = cens_si_93cn, dist = "W", optim.method="SANN",
   seed=1234)
si_cens_W = coarse2estim(censans_W, thin=10)
save(si_cens_W, file="si_cens_W.rda")
censans_L = coarseDataTools::dic.fit.mcmc(dat = cens_si_93cn, dist = "L", optim.method="SANN",
   seed=1234)
si_cens_L = coarse2estim(censans_L, thin=10)
save(si_cens_L, file="si_cens_L.rda")
censans_G = coarseDataTools::dic.fit.mcmc(dat = cens_si_93cn, dist = "G", optim.method="SANN",
   seed=1234)
si_cens_G = coarse2estim(censans_G, thin=10)
save(si_cens_G, file="si_cens_G.rda")
```

On the left we trace the pointwise means of the posterior distribution of SI(t).
On the right we give the posterior density of SI(4) for three
parametric models used to form the censored data likelihoods.

```{r lkpl, fig.height=4, fig.width=7, echo=FALSE}
ln = si_cens_L
w = si_cens_W
g = si_cens_G
par(mfrow=c(1,2))
plot(1:40,apply(ln[[1]], 1, mean)[1:40], type="l", lwd=2, xlab="days from origin", ylab="density",
    col="darkgray", main="SI(t), posterior mean (pointwise)")
lines(1:40, apply(w[[1]], 1, mean)[1:40], type="l", lty=2, lwd=2, col="darkgray")
lines(1:40, apply(g[[1]], 1, mean)[1:40], type="l", lty=3, lwd=2, col="darkgray")
legend(8, .14, lty=1:3, legend=c("lognormal", "Weibull", "Gamma"), col="darkgray", bty="n", lwd=2)

plot(density(ln[[1]][4,]), xlim=c(.10,.18), lwd=2, col="darkgray", 
    main="Posterior mass at day 4", xlab="MCMC mass", ylim=c(0,90))
lines(density(w[[1]][4,]), lwd=2, col="darkgray", lty=2)
lines(density(g[[1]][4,]), lwd=2, col="darkgray", lty=3)
```

### Using only positive values for transmission time

```{r dopos, eval=FALSE, echo=FALSE, message=FALSE}
posans_W = coarseDataTools::dic.fit.mcmc(dat = pos_si_93cn, dist = "W", optim.method="SANN",
   seed=1234, verbose=0)
si_pos_W = coarse2estim(posans_W, thin=10)
save(si_pos_W, file="si_pos_W.rda")
posans_L = coarseDataTools::dic.fit.mcmc(dat = pos_si_93cn, dist = "L", optim.method="SANN",
   seed=1234, verbose=0)
si_pos_L = coarse2estim(posans_L, thin=10)
save(si_pos_L, file="si_pos_L.rda")
posans_G = coarseDataTools::dic.fit.mcmc(dat = pos_si_93cn, dist = "G", optim.method="SANN",
   seed=1234, verbose=0)
si_pos_G = coarse2estim(posans_G, thin=10)
save(si_pos_G, file="si_pos_G.rda")
```

```{r lkpl2, fig.height=4, fig.width=7, echo=FALSE}
ln = si_pos_L
w = si_pos_W
g = si_pos_G
par(mfrow=c(1,2))
plot(1:40,apply(ln[[1]], 1, mean)[1:40], type="l", lwd=2, xlab="days from origin", ylab="density",
    col="darkgray", main="SI(t), posterior mean (pointwise)")
lines(1:40, apply(w[[1]], 1, mean)[1:40], type="l", lty=2, lwd=2, col="darkgray")
lines(1:40, apply(g[[1]], 1, mean)[1:40], type="l", lty=3, lwd=2, col="darkgray")
legend(8, .14, lty=1:3, legend=c("lognormal", "Weibull", "Gamma"), col="darkgray", bty="n", lwd=2)

plot(density(ln[[1]][4,]), xlim=c(.10,.18), lwd=2, col="darkgray", 
    main="Posterior mass at day 4", xlab="MCMC mass", ylim=c(0,90))
lines(density(w[[1]][4,]), lwd=2, col="darkgray", lty=2)
lines(density(g[[1]][4,]), lwd=2, col="darkgray", lty=3)
```

# Estimating R[t] using stored MCMC-based fits to SI data

## Using the censored representation of SI data

```{r getdat, cache=FALSE}
suppressWarnings({dat = enriched_jhu_data()})
cusa = cumulative_events_ejhu(dat, eventtype="confirmed",
      alpha3="USA")
cusa
iusa = form_incident_events(trim_from(cusa, "2020-03-10"))
iusa2 = data.frame(I=iusa$count, dates=iusa$dates)
```

### Weibull
```{r fit1, fig.height=6, cache=TRUE}
wcens = EpiEstim::estimate_R(iusa2, method="si_from_sample", si_sample=si_cens_W$si_sample, 
     config=make_config(list(n2=50, seed=1)))
nn = plot(wcens)
```

### Lognormal

```{r fit2, fig.height=6, cache=TRUE}
lncens = EpiEstim::estimate_R(iusa2, method="si_from_sample", si_sample=si_cens_L$si_sample, 
     config=make_config(list(n2=50, seed=1)))
nn = plot(lncens)
```

### Comparing R[t] trajectories for Weibull and Lognormal base models

```{r senso}
plot(wcens$dates[-c(1:7)], wcens$R$`Mean(R)`, xlab="date", ylab="R(t)", type="l")
lines(lncens$dates[-c(1:7)], lncens$R$`Mean(R)`,lty=2)
legend("topright",lty=c(1,2),legend=c("Weibull", "lognormal"))
```

## Using the positive-SI only representations and fits

### Weibull

```{r fit1b, fig.height=6, cache=TRUE}
wpos = EpiEstim::estimate_R(iusa2, method="si_from_sample", si_sample=si_pos_W$si_sample, 
     config=make_config(list(n2=50, seed=1)))
nn = plot(wpos)
```

### Lognormal

```{r fit2b, fig.height=6, cache=TRUE}
lnpos = EpiEstim::estimate_R(iusa2, method="si_from_sample", si_sample=si_pos_L$si_sample, 
     config=make_config(list(n2=50, seed=1)))
nn = plot(lnpos)
```

### R[t] trajectories for Weibull and Lognormal base models, censored and positive-only

```{r sensocomp}
plot(wcens$dates[-c(1:7)], wcens$R$`Mean(R)`, xlab="date", ylab="R(t)", type="l")
lines(lncens$dates[-c(1:7)], lncens$R$`Mean(R)`,lty=2)
lines(wpos$dates[-c(1:7)],wpos$R$`Mean(R)`,lty=3)
lines(lnpos$dates[-c(1:7)], lnpos$R$`Mean(R)`,lty=4)
legend("topright",lty=c(1,2),legend=c("Weibull", "lognormal"))
```

We conclude that the difference between excluding negative SI reports and
censoring the infector times is very small.  Further work is needed to employ
all the information (i.e., the negative SI values) provided in the Du et al. CDC report.
