---
title: "sars2pack: packaging of C. Morefield code on COVID-19 epidemic modeling"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{sars2pack: packaging of C. Morefield code on COVID-19 epidemic modeling}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

```{r setup,echo=FALSE}
suppressMessages({
suppressPackageStartupMessages({
library(BiocStyle)
library(sars2pack)
library(dplyr)
library(tibble)
library(lubridate)
})
})
```

# Introduction

John C Mallery conveyed code of Charles Morefield to harvest
COVID-19 time series data.  Vince Carey has started an R package/github repo
to manage the relevant code.  The package is called sars2pack in hopes of
avoiding name conflict with many other packages likely developing.

The main functionalities thus far concern importing data from
Johns Hopkins Whiting School github repo `github.com/CSSEGISandData` and
using the `r CRANpkg("R0")` package to perform certain tasks of
epidemic modeling.  We also retain a snapshot of existing time series
data from 19 March 2020 to facilitate demonstration.

# Data and updating

## Access

The snapshot looks like:
```{r lkda}
as_tibble(mar19df)
```

We can call `fetch_JHU()_Data` to get a more current instance.

## Structure

There are nested structures in the data.  Provinces or states nest within countries.

```{r lkcou}
head(sort(table(mar19df$CountryRegion),decreasing=TRUE))
```

```{r nesting}
mar19df %>% select(CountryRegion, ProvinceState) %>% 
   filter(CountryRegion=="France")
```

The code to assemble information for a country, out of its constituent records,
needs to be looked at.  The Morefield code for extraction uses data.table idioms -- I am
not completely sure it is correct.  I would like our extracts to be precisely formulated.

Looking at the raw data seems important to get a clear sense
of the modeling challenge.
```{r doser,fig.height=7}
get_series = function(province="", country, dataset=sars2pack::mar19df) {
  ans = dataset %>% filter(ProvinceState==province & CountryRegion==country)
  ans[,-c(1:4)]
}
plot_series = function(province="", country, dataset=sars2pack::mar19df, ...) {
 ser = get_series(province=province, country=country, dataset=dataset)
 dates = lubridate::as_date(mdy(names(dataset)[-c(1:4)]))
 plot(dates, ser, main=paste(province, country), ...)
}
par(mfrow=c(2,2),mar=c(4,3,2,2))
plot_series(province="Hubei", country="China")
plot_series(country="Italy")
plot_series(country="Thailand")
plot_series("Massachusetts", country="US")
```


# Modeling illustrated for a simulation

Following code conveyed by John Mallery, we have the following
approach for estimating R0 using a single realization of
an epidemic simulation.

Note that there can be failures of `estimate.R` for certain
inputs.  We are working on that.
```{r dostraight}
library(R0)
# Generating an epidemic with given parameters
mGT <- generation.time("gamma", c(3,1.5))
set.seed(5432)  # always initialize when simulating!
mEpid <- sim.epid(epid.nb=1, GT=mGT, epid.length=30, 
     family="poisson", R0=1.67, peak.value=500)
mEpid <- mEpid[,1]
# Running estimations
est <- estimate.R(epid=mEpid, GT=mGT, methods=c("EG","ML","TD"), begin=1, end=30)
```

We modified the plotting function in `r CRANpkg("R0")` which
was calling `dev.new` too often.  Use `plot2`.

```{r lksim,fig.height=7}
par(mfrow=c(2,2))
plot2(est)
```

The plotfit2 function is also useful.  These fits
look identical but they are not.

```{r lksim2, fig.height=7}
par(mfrow=c(2,2))
plotfit2(est)
```

# Modeling for a geographic entity

Now we extract information from the time-series table and
obtain estimates of R0 under exponential growth.

## Hubei Province

We are able to use exponential growth and time-dependent models
with this data, using generation time model from a
recent [Annals of Internal Medicine](https://annals.org/aim/fullarticle/2762808/incubation-period-coronavirus-disease-2019-covid-19-from-publicly-reported) paper.

The incidence data probably need smoothing, and the time-dependent
model has unreasonable fluctuations.

```{r dohub,fig.height=7}
dates = lubridate::as_date(mdy(names(mar19df)[-c(1:4)]))
hubdat = as.numeric(get_series(province="Hubei", country="China", 
    dataset=sars2pack::mar19df))
names(hubdat) = dates
mGT <- generation.time("gamma", c(5.8, 0.95)) # from DOI 10.7326/M20-0504
hubdat.filt = trimLeading(c(hubdat[1], diff(hubdat)))
est.EG <- estimate.R(epid=hubdat.filt, GT=mGT, 
    methods=c("EG", "TD"), begin=1L, end=as.integer(length(hubdat.filt)))
est.EG
par(mfrow=c(2,2), mar=c(5,3,2,2))
plot2(est.EG)
plotfit2(est.EG)
```

## Italy

For Italy, only the EG model seems to work, with the
Annals of Internal Medicine generation time model.  It
fits the data reasonably well, but the data seems to include
a reporting gap.

```{r doit,fig.height=7}
itdat = as.numeric(get_series(province="", country="Italy", 
    dataset=sars2pack::mar19df))
names(itdat) = dates
itdat.filt = trimLeading(c(itdat[1], diff(itdat)))
est.EG <- estimate.R(epid=itdat.filt, GT=mGT, 
    methods=c("EG"), begin=1L, end=as.integer(length(itdat.filt)))
est.EG
par(mfrow=c(2,2), mar=c(5,3,2,2))
plot2(est.EG, main="Italy")
plotfit2(est.EG, main="Italy")
```
